---
title: "exercise-12"
author: "Riley N Derby"
date: "2024-04-09"
output: html_document
---


```{r}
library(tidyverse)
library(dplyr)
```

read in mammal dataset
```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt"
d <- read_tsv(f, col_names = TRUE)
glimpse(d)
str(d)
```


step 1: Replace all values of -999 (the authors’ code for missing data) with NA.
```{r}
d <- mutate_all(d, ~ifelse(. == -999, NA, .))
```

step 2: Drop the variables litter size and refs.
```{r}
d <- d %>%
  select(-c('litter size', refs))
str(d)
```

step 3:Log transform all of the other numeric variables.

```{r}
d <- d %>% mutate(
  LogMass = log(`mass(g)`),
  LogGestation = log(`gestation(mo)`),
  LogNewborn = log(`newborn(g)`),
  LogWeaning= log(`weaning(mo)`),
  LogWeanMass = log(`wean mass(g)`),
  LogAFR= log(`AFR(mo)`),
  LogMaxLife= log(`max. life(mo)`))


```

step 4: Regress the (now log transformed) age [gestation(mo), weaning(mo), AFR(mo) (age at first reproduction), and max. life(mo) (maximum lifespan)] and mass [newborn(g) and wean mass(g)] variables on (now log transformed) overall body mass(g) and add the residuals to the dataframe as new variables (relGest, relWean, relAFR, relLife, relNewbornMass, and relWeaningMass).

```{r}
m <- lm(data = d, LogMass ~ LogGestation + LogNewborn + LogWeaning + LogWeanMass + LogAFR + LogMaxLife, na.action = na.exclude)
summary(m)

relGest <- resid(lm(LogMass ~ LogGestation, data = d, na.action = na.exclude))
relWean <- resid(lm(LogMass ~ LogWeaning, data = d, na.action = na.exclude))
relAFR <- resid(lm(LogMass ~ LogAFR, data = d, na.action = na.exclude ))
relLife <- resid(lm(LogMass ~ LogMaxLife, data = d, na.action = na.exclude))
relNewbornMass <- resid(lm(LogMass ~ LogNewborn, data = d, na.action = na.exclude))
relWeaningMass <- resid(lm(LogMass ~ LogWeanMass, data = d, na.action = na.exclude))

d$relGest <- relGest
d$relWean <- relWean
d$relAFR <- relAFR
d$relLife <- relLife
d$relNewbornMass <- relNewbornMass
d$relWeaningMass <- relWeaningMass
```

step 5: Plot residuals of max lifespan (relLife) in relation to Order. Which mammalian orders have the highest residual lifespan?
```{r}
## sometimes, I like flipping the axes to get a better look... which is what i did here
ggplot(data = d, aes(x = relLife, y = order)) +
  geom_boxplot() +
  theme_minimal() 
```
Sirenia and Cetacea have highest residual lifespan



Plot residuals of newborn mass (relNewbornMass) in relation to Order. Which mammalian orders have the have highest residual newborn mass?
```{r}
ggplot(data = d, aes(x = relNewbornMass, y = order)) +
  geom_boxplot() +
  theme_minimal() 
```
Carnivora have highest residual newborn mass


Plot residuals of weaning mass (relWeaningMass) in relation to Order. Which mammalian orders have the have highest relative weaning mass?
```{r}
ggplot(data = d, aes(x = relWeaningMass, y = order)) +
  geom_boxplot() +
  theme_minimal() 
```
Philodota and Lagomorpha have some of the highest, but there are some very high outliars in Carnivora and Rodentia



step 6: Run models and a model selection process to evaluate what (now log transformed) variables best predict each of the two response variables, max. life(mo) and AFR(mo), from the set of the following predictors: gestation(mo), newborn(g), weaning(mo), wean mass(g), litters/year, and overall body mass(g).

lets do max lifespan model first
```{r}
d_life <- d %>%
  drop_na(LogMaxLife, LogMass, LogGestation, LogNewborn, LogWeaning, LogWeanMass, `litters/year`)
```


```{r}
#start with

m_null <- lm(data = d_life, LogMaxLife ~ 1)


##add in highest F value to new update each time
add1(m_null, scope = .~. + LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, test = "F")
m1 <- update(m_null, formula = .~. + LogGestation)
summary(m1)

add1(m1, scope = .~. + LogMass + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, test = "F")
m2<- update(m1, formula = .~. + `litters/year`)
summary(m2)

add1(m2, scope = .~. + LogMass + LogNewborn + LogWeaning + LogWeanMass , test = "F")
m3<- update(m2, formula = .~. + LogMass)
summary(m3)

add1(m3, scope = .~. + LogNewborn + LogWeaning + LogWeanMass, test = "F")
m4<- update(m3, formula = .~. + LogWeaning)
summary(m4)

```
lifespan is best predicted by 1.gestation length, 2.litters per year, 3.mass, and then 4. weaning age


now AFR
```{r}
d_AFR <- d %>%
  drop_na(LogAFR, LogMass, LogGestation, LogNewborn, LogWeaning, LogWeanMass, `litters/year`)

#start with

m_null2 <- lm(data = d_AFR, LogAFR ~ 1)


##add in highest F value to new update each time
add1(m_null2, scope = .~. + LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, test = "F")
m1 <- update(m_null, formula = .~. + LogGestation)
summary(m1)

add1(m1, scope = .~. + LogMass + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, test = "F")
m2<- update(m1, formula = .~. + `litters/year`)
summary(m2)

add1(m2, scope = .~. + LogMass + LogNewborn + LogWeaning + LogWeanMass , test = "F")
m3<- update(m2, formula = .~. + LogWeaning)
summary(m3)

add1(m3, scope = .~. + LogNewborn + LogMass + LogWeanMass, test = "F")
m4<- update(m3, formula = .~. + LogMass)
summary(m4)

```
the predictors of age at first reproduction are 1. gestation length, 2. litters per year, 3. weaning age, 4. mass

For each of the two response variables, indicate what is the best model overall based on AICc and how many models have a delta AICc of 4 or less.

```{r}
library(MASS)
m_full <- lm(LogMaxLife ~ LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, data = d_life)
s <- stepAIC(m_full, scope = .~., direction = 'both', trace = T)


m_full2 <- lm(LogAFR ~ LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, data = d_AFR)
s2 <- stepAIC(m_full2, scope = .~., direction = 'both', trace = T)
```
What variables, if any, appear in all of this set of “top” models?

For max lifespan, the best model includes logmass, loggestation, logweaning, and litters per year

For age first reproduction the best model includes the same 4 predictors


Calculate and plot the model-averaged coefficients and their CIs across this set of top models.

```{r}
library(MuMIn)
m_full <- lm(LogMaxLife ~ LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, data = d_life, na.action = na.fail)

mods <- dredge(m_full)

mods.res <- get.models(mods, subset = T)
mods.res$'24'

mods.res <- get.models(mods, subset = delta <= 4) #returns top models where delta aicc <= 4

mods.avg <- summary(model.avg(mods, subset = delta <= 4, fit = T))# averages coeff for top models where delta.aicc <=4


mods.avg <- summary(model.avg(mods, subset = cumsum(weight) <= 0.99, fit = T))# cumsum of weights of all models will eqyal 1.... averages coeff for top models where cumulative weight is <=0.99

confint(mods.avg)

plot(mods.avg, full = T)
plot(mods.avg, full = F)

#####afr model
m_full2 <- lm(LogAFR ~ LogMass + LogGestation + LogNewborn + LogWeaning + LogWeanMass + `litters/year`, data = d_AFR, na.action = na.fail)

mods2 <- dredge(m_full2)

mods.res2 <- get.models(mods2, subset = T)
mods.res2$'24'

mods.res2 <- get.models(mods2, subset = delta <= 4) #returns top models where delta aicc <= 4

mods.avg2 <- summary(model.avg(mods2, subset = delta <= 4, fit = T))# averages coeff for top models where delta.aicc <=4


mods.avg2 <- summary(model.avg(mods2, subset = cumsum(weight) <= 0.99, fit = T))# cumsum of weights of all models will eqyal 1.... averages coeff for top models where cumulative weight is <=0.99

confint(mods.avg2)

plot(mods.avg2, full = T)
plot(mods.avg2, full = F)
```

for lifespan model, there are 4 models with AIC less than 4
for AFR model, there are 5 models with AIC less than 4


